---
title: "Diagnostics for n-way median polish"
output: html_document
date: "2026-05-26"
editor_options: 
  chunk_output_type: console
---

```{r colored_table ,include=FALSE}
colored_table <- function(df, palette = "Greens", n_quantiles = 5,
                          digits = 2, border = "grey50", range = c(9.25,19),
                          nbrks = 20,
                          x_axis=FALSE, y_axis=FALSE) {
  
  # Convert to matrix
  mat <- as.matrix(df)
  
  # Flatten values and remove NA for quantiles
  vals <- as.numeric(mat)
  vals_no_na <- vals[!is.na(vals)]
  
  # Compute quantile breaks
  # breaks <- quantile(vals_no_na,
  #                    probs = seq(0, 1, length.out = n_quantiles + 1),
  #                    na.rm = TRUE)
  breaks <- seq(range[1],range[2], length.out=nbrks)
  
  # Ensure unique breaks (important if many duplicates)
  breaks <- unique(breaks)
  
  # Create color palette (requires grDevices, which is base)
  # cols <- colorRampPalette(
  #   c("#edf8e9", "#bae4b3", "#74c476", "#31a354", "#006d2c")
  # )(length(breaks) - 1)
  cols <- hcl.colors(nbrks -1 , "Greens", rev = TRUE)
  
  # Assign colors to each value
  bin_index <- cut(mat, breaks = breaks, include.lowest = TRUE, labels = FALSE)
  cell_cols <- matrix(cols[bin_index], nrow = nrow(mat))
  
  # Plot setup
  nr <- nrow(mat)
  nc <- ncol(mat)
  
  OP2 <- par(mar = c(3,5,1,1))
  plot(0, 0, type = "n",
       xlim = c(0, nc), ylim = c(0, nr),
       xaxt = "n", yaxt = "n", xlab = "", ylab = "",
       bty = "n")
  
  # Draw cells
  for (i in 1:nr) {
    for (j in 1:nc) {
      rect(j - 1, nr - i,
           j, nr - i + 1,
           col = cell_cols[i, j],
           border = border)
      
      # Add text
      txt <- ifelse(is.na(mat[i, j]), "", round(mat[i, j], digits))
      text(j - 0.5, nr - i + 0.5, labels = txt, cex = 0.8)
    }
  }
  
  # Add row/column labels
  OP1 <- par(family = "mono")
   if(!identical(x_axis, FALSE))  axis(1, at = seq(0.5, nc - 0.5), labels = x_axis, 
                                       tick = FALSE, line = -1, cex.axis = 0.8)
   if(!identical(y_axis, FALSE))  axis(2, at = seq(nr - 0.5, 0.5, -1), labels = y_axis, 
                                       tick = FALSE, line = -1, las=1, cex.axis = 0.8)
  par(OP1)
par(OP2)
}


```

# PART 1: How ANOVA "Sees" Effects

## Datasets used in this section

We'll start with a simple example where the response variable is tomato **yield** (`Y`) in pounds. Let's assume that an experiment is conducted whereby Nitrogen (`N`) and Phosphorous (`P`)--two important plant nutrients--are used to fertilize plants at different levels of concentration. 

### No fertilizer effect 

Let's start off with a scenario where we have six tomato plants that are not given fertilizer. In such a scenario the tomato plants produce an average of around 10 lbs of tomatoes. We'll assume for now that there is no random variability in plant yield (i.e. each produces exactly 10 pounds of tomatoes).

```{r echo=FALSE, fig.height=2.1, fig.width = 2.5}
library(tukeyedar)
base <-rep(10,6)
df   <- data.frame(base, Plevel = rep(c("P(0.5)","P(1.5)","P(2.0)"),2), P = rep(c(0.5,1.5,2),2), 
                         Nlevel=  rep(c("N(1)","N(2)"),3), N = rep(c(1,2),3))
df$yield <- df$base
eda_matlong(df, yield, P, N, direction = "to_array") |>  colored_table()
```

This will be the baseline model. At this point, there are no `N` or `P` effects given that we did not incorporate them in the potted plants.

### Add a Phosphorous effect

Next, we'll include our first effect: We'll add different Phosphorous concentrations (`P`) across rows so that each row of plants that are laid out in the matrix is assigned the same concentration. We'll add three different concentrations that we'll name `P0.5`, `P1.5` and `P2.0`. Each concentration increases tomato yield by 0.5, 1.5 and 2 pounds respectively. 

This gives us the following model:

$$
Yield_{(P,\ N)} = 10 + 0.5(P0.5) + 1.5(P1.5) + 2(P2.0)
$$
Where the `P0.5`, `P1.5` and `P2.0` are binary values that take on a value of `0` or `1` depending on which Phosphorous treatment was applied to the potted plants.

This augments our tomato yield as follows:

```{r echo=FALSE, fig.height=2.1, fig.width = 2.5}
library(tukeyedar)
df$yieldP <- df$base + df$P
eda_matlong(df, yieldP, P, N, direction = "to_array") |>  colored_table(y_axis = paste0(unique(df$Plevel), "->"))
```


### What an ANOVA gives us

An ANOVA analysis does not only perform a hypothesis test (outcome that results in a "p-value"). An ANOVA also provides us with the contribution each factor (aka category) makes in the response variable (tomato yield in this example). 

Using the base R environment, one may extract each factor's **effect** using the command:

```{r}
aov(yieldP ~ Plevel, df) |> model.tables()
```

This above approach makes use of least square method--a popular implementation of an ANOVA analysis.

The `aov()` function is the standard, but for the rest of this article, we'll use a function from the `tukeyedar` package called `eda_mean_sweep()`. It   calculates the exact same effects, but presents them in a way that will be very helpful as we build more complex models.

```{r}
MP <- eda_mean_sweep(df, yieldP, P) 
MP$effects
```

You'll notice that the effect values for `Plevel` do not reflect the values used in our initial model. In other words, instead of seeing the modeled values of [`r df$P[1:3]`], we have effect values of [`r round(MP$effects$P, 3)`]. 

You will also note that the overall mean value (also known as the **grand mean** or the **global value**) is not 10 but instead is,

```{r}
MP$global
```

This results in a model of:

$$
Yield = 11.334 -0.833(P0.5) + 0.167(P1.5) + 0.667(P2.0) 

$$
This might look like a completely different model, but it's just a different way of telling the same story. Let's calculate the yield for a plant given the `P1.5` treatment:

$$
Yield_{(P1.5,\ N2)} = 11.334 -0.833(0) +0.167(1) + 0.667(0) = 11.5 
$$
And where have we seen that number before? It's the exact same yield we   created with our simple "building block" model $10 + 1.5 = 11.5$. The models are different, but the result is identical.

 So why the different numbers? Because an analysis of variance doesn't try to find the original "building block" recipe. Instead, it describes the world in terms of **deviations from an average**, telling us how much *better* or *worse* each factor (or treatment in our example) is compared to the overall mean of the experiment.

In out working example, the model states that the Phosphorous treatment of 1.5 units of Phosphorous (`P1.5`) adds a yield of `0.167` above an overall yield of `11.334` in this experiment. Likewise, it also states that a Phosphorous treatment of 0.5 units of Phosphorous (`P0.5`) generates a yield that is `0.833` **less** than the overall mean in this experiment.

## A two-way analysis

### A purely additive effect

So what happens when we add Nitrogen as a treatment to our tomato plants? One possible outcome is that the effect is solely additive meaning that the benefits afforded by each nutrient are added to one another. The previous model can be augmented as,

$$
Yield_{(P,\ N)} = 10 + 0.5(P0.5) + 1.5(P1.5) + 2(P2.0) + 1(N1) + 2(N2)
$$
As with the Phosphorous factors, the Nitrogen factors take on a value of `0` or `1` depending on which column of tomatoes is assigned the treatment. 

```{r echo=FALSE, fig.height=2.1, fig.width = 2.5}
df$yieldPN <- df$base + df$N + df$P
eda_matlong(df, yieldPN, P, N, direction = "to_array") |>  
  colored_table(x_axis = paste0(unique(df$Nlevel), "🡅"),
                y_axis = paste0(unique(df$Plevel), "->"))
```

Now that we've created a dataset with two independent, additive effects, let's see what the ANOVA "report" looks like. We'll run a two-way analysis of variance and look at the effects it extracts.

```{r}
MPN <- eda_mean_sweep(df, yieldPN, P, N)
MPN
```

The output gives us a gloabl mean and the effects for both `Plevel` and `Nlevel.`

   * Grand Mean: `r round(MPN$global, 3)` 
   * `Plevel` effects: `r round(MPN$effects$P, 3)` 
   * `Nlevel` effects: `r round(MPN$effects$N, 3)` 

This is the classic signature of an additive model. The term "additive" means the effect of one nutrient **doesn't change based on the level of the other**. The *bonus* from using Nitrogen is the same regardless of the Phosphorous treatment, and vice-versa. Because we created the data this way, the ANOVA model correctly shows two separate, independent sets of effects.


The ANOVA model can thus be expressed as:

$$
Yield_{(P,\ N)} = 12.83 - 0.833(P0.5) + 0.167(P1.5) + 0.667(2.0) - 0.5(N1) + 0.5(N2)
$$
Here to, this is nothing more than a re-expression of the base model used to generate the data. To confirm this, we can compute the estimated tomato yield for a plant assigned a Phosphorous treatment of `P1.5` and a Nitrogen treatment of `N2.0`:

$$
Yield_{(P1.5,\ N2)} = 12.83 - 0.833(0) + 0.167(1) + 0.667(0) - 0.5(0) + 0.5(1) = 13.5
$$
This is the same yield generated from the base model.

The ANOVA output correctly suggests that the `P` treatment effect is greater than the `N` treatment effect meaning that the Phosphorous factor, as a whole, creates more variation in the yield than the Nitrogen factor does. This doesn't mean that every level of P is better than N, but that the choice of Phosphorous level has a bigger impact on the final outcome. 

Here's how we can tell from the effect values:
   * For Nitrogen, the effects are `-0.5` and `+0.5`. The "swing" from the worst
     level (N1) to the best level (N2) is `(+0.5) - (-0.5) = 1` pound of
     tomatoes.
   * For Phosphorous, the effects range from `-0.833` to `+0.667`. The "swing"
     from the worst level (P0.5) to the best level `(P2.0) is (+0.667) -
     (-0.833) = 1.5` pounds.

Because the total range of influence is bigger for Phosphorous (a 1.5 lb swing) than for Nitrogen (a 1.0 lb swing), the ANOVA correctly identifies that P is the more impactful factor in our experiment for explaining the variation in tomato yield.

Calculating the range of effects is one way to see which factor is more impactful. An even more direct way is to visualize the effects on a common scale using a **variability decomposition plot**.

```{r fig.height = 3.5, fig.width=4}
plot(MPN, show.resp = TRUE, label = TRUE)
``` 

This plot perfectly summarizes our analysis. Let's break it down:


   * On the far left, the **Response** boxplot shows the total spread of our final
     tomato yields (`yieldPN`), centered around the global mean. This represents
     the total variation we are trying to explain.
   * The next two sets of points show the effects for `Plevel` and `Nlevel.` Each
     dot represents the "nudge" (positive or negative) that a specific
     treatment level gives to the yield.

Most importantly, you can now see which factor is more powerful. The spread of the dots for Plevel is visibly wider than the spread for Nlevel. This is the graphical proof of what we calculated manually: the 1.5 lb swing from Phosphorous has a larger impact than the 1.0 lb swing from Nitrogen.

You will also notice a component on the plot labeled Residuals which is   empty (all values are zero). This is because the Residuals represent the "leftover" variation that our model **cannot explain**. Since we created a perfect, "textbook" dataset with no random variability, our additive model of P and N effects accounts for 100% of the variation in the yield. This leaves nothing behind. We can confirm this by extracting the residuals from the `MPN` model.

```{r}
MPN$residuals
```

Note that whenever you see a number ending in `e-15` or `e-16`, you can confidently treat it as the computer's way of saying "zero."
  
Later, when we introduce natural variability, this residuals component will be crucial for understanding our model's fit.


### Adding an interactive effect

An interaction means the effect of one factor depends on the level of another. Let's create a classical ANOVA interaction. We'll model a synergistic effect where the combination of the highest levels of Nitrogen (N2) and Phosphorous (P3) gives a special boost to the yield that is more than the sum of its parts.

We'll create an explicit interaction term. It will be 0 for all combinations except for N=2 and P=2, where we'll add a boost of 2.5.


```{r echo=FALSE, fig.height=2.1, fig.width = 2.5}
df$interaction <- ifelse(df$N == 2 & df$P == 2, 2.5, 0)
df$yieldPN2 <- df$base + df$N + df$P + + df$interaction
eda_matlong(df, yieldPN2, P, N, direction = "to_array") |>  
  colored_table(x_axis = paste0(unique(df$Nlevel), "🡅"),
                y_axis = paste0(unique(df$Plevel), "->"))
```

Now, if we were to fit the additive model used earlier, we would not be able to fully capture the interactive effect.

```{r}
MPN2 <- eda_mean_sweep(df, yieldPN2, P, N)
MPN2$effects
MPN2$global
```

We have a model of the form,

$$
Yield_{(P,\ N)} = 13.25 - 1.25(P0.5) -0.25(P1.5) + 1.5(P2.0) - 0.92(N1) + 0.92(N2)
$$
If we use this additive model to reconstruct the tomato yield for the `P2.0:N2` treatment combination, we get the following yield

$$
Yield_{(P2.0,\ N2)} = 13.25 - 1.25(0) -0.25(0) + 1.5(1) - 0.92(0) + 0.92(1) = 15.67
$$
This computed yield is nearly one pound less than the expected yield of `r df$yieldPN2[6]`.

If an interaction between effects is suspected, than the interactive component needs to be added to the model. With the `eda_mean_sweep` function, we add the interactive component via the `max_order` component. We set this value to `2` indicating that we are seeking all two-way interactions.

```{r}
MPN3 <- eda_mean_sweep(df, yieldPN2, P, N, max_order = 2)
MPN3$effects
MPN3$global
```

Adding the interactive component (`max_order = 2`) creates a more complex model. It still has the main effects for P and N, but it adds a specific adjustment value for each unique P:N combination. These adjustments are the interaction effects, and they account for the synergy that the simple additive model missed. We now have a (correct) model of the form,

$$
\begin{align}
{Yield_{(P,\ N)}} &=  13.25 - 1.25(P0.5) -0.25(P1.5) + 1.5(P2.0) - 0.92(N1) + 0.92(N2) \\
             &0.417(P0.5:N1) + 0.417(P1.5:N1) - 0.833(P2.0:N1) \\
             &- 0.417(P0.5:N2) - 0.417(P1.5:N2) + 0.833(P2.0:N2)
\end{align}
$$
Where the interactive factor levels (e.g. `(P0.5:N1)`) are binary values that take on a `1` if the response value is associated with that combination of effects.

Recall from our earlier discussion that the ANOVA does not seek to recreated the base model used to generate the data. As such, it's no surprise that non-zero values are associated with the interactive effects that were explicitly set to 0 in the base model.

Revisiting our example of the tomato yield for a treatment of `P2.0:N2`, we can now use the full model, which includes the main effects plus the specific interaction effect for that cell

$$
\begin{align}
{Yield_{(P,\ N)}} &=  13.25 - 1.25(0) -0.25(0) + 1.5(1) - 0.92(0) + 0.92(1) \\
             &0.417(0) + 0.417(0) - 0.833(0) \\
             &- 0.417(0) - 0.417(0) + 0.833(1) = 16.5
\end{align}
$$
This matches the `r df$yieldPN3[6]` lbs yield we originally generated. By including the interaction term, the model can now perfectly explain the data.

### Residuals

So, what does this failure of the additive model (`MPN2`) look like visually? Let's look at the decomposition plot for the additive-only model that we fitted to our interactive data.

```{r fig.height = 3.5, fig.width=4}
plot(MPN2, show.resp=TRUE, label = TRUE)
```

This plot tells a clear story. Previously, when we analyzed a truly additive dataset, the **Residuals** component was empty. But look at it now! It has a very wide spread, with values ranging from about `-0.833` to `+0.833`.

This is the visual proof that our model is wrong. Those large residuals are   the *"synergy"* from the interaction that the simple additive model could not explain.  In statistics, we give this leftover, unexplained variation a formal name: the **Residual**, often symbolized by the Greek letter $\varepsilon$.  

As such, when fitting a model to data other than one created synthetically (as is the case here), you want to include a **Residuals** component, $\varepsilon$ to your model. Given that some of the yield data in `df$yieldPN2` was not captured by the `N` and `P` components, our earlier equation was not complete. Adding $\varepsilon$ completes the model.

$$
Yield_{(P,\ N)} = 13.25 - 1.25(P0.5) -0.25(P1.5) + 1.5(P2.0) - 0.92(N1) + 0.92(N2) +\varepsilon 
$$
When we tried reconstructing the yield (`df$yieldPN2`) for the `P2.0:N2` combination, we were short 0.833 lbs. That value was, in fact, the residual $\varepsilon$ for that specific observation.

By explicitly adding the interaction effect to our model, we give that *leftover* variation a proper home. The variation that was previously  unexplained and forced into the $\varepsilon$ term is now correctly captured   by the `P:N` interaction term.

```{r fig.height = 3.5, fig.width=5}
plot(MPN3, show.resp=TRUE, label = TRUE)
```

The P:N interaction term now appears in the plot, and it has captured all the variation that was previously "leaking" into the residuals. The *Residuals* component is now empty again (because we did not include unexplained variability in the base model), telling us that our full model perfectly explains the data we created.




## Add natural (random) variability

Let's start off with a scenario where we have six tomato plants that are not given fertilizer. In such a scenario the tomato plants produce an average of around 10 lbs of tomatoes. This is the **baseline** yield. Given natural variability, the six plants may not generate exactly 10 lbs worth of tomatoes, for example, they may generate `9.5`, `10.5`, `9`, `9.5`, `10`, `11` pounds respectively. 

```{r  echo=FALSE}
base <- c(9.5, 9.5, 10.25, 10, 11.25, 11.25)
df   <- data.frame(base, Plevel = rep(c("P(0.5)","P(1.5)","P(2.0)"),2), P = rep(c(0.5,1.5,2),2), 
                         Nlevel=  rep(c("N(1)","N(2)"),3), N = rep(c(1,2),3))
```

If we layout the plants in a grid, we have the following yield distributions:

```{r echo=FALSE}

df$yield <- df$base

eda_matlong(df, yield, P, N, direction = "to_array") |>  
  colored_table(x_axis = paste0(unique(df$Nlevel), "🡅"),
                y_axis = paste0(unique(df$Plevel), "->"))
```








