---
title: "The residual-fit spread plot"
author: "Manuel Gimond"
date: "`r Sys.Date()`"
output: 
  rmarkdown::html_vignette:
    toc: TRUE
    css: style.css
vignette: >
  %\VignetteIndexEntry{The empirical QQ plot (and derived Tukey mean-difference plot)}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#> ",
  message = FALSE,
  tidy = FALSE,
  cache = FALSE,
  warning = FALSE,
  encoding = "UTF-8"
)
```

```{r include=FALSE}
knitr::opts_chunk$set(echo = TRUE, dev.args=list(pointsize=10))
```

# Introduction

The *residual-fit spread plot* (rfs plot) is a plot first introduced by William Cleveland designed to compare the variability explained by a fitted model (such as group means for a univariate analysis or a linear model for a bivariate analysis) and the variability in the residuals. It's constructed as follows:

-  A model is fitted to the data.
-  The residuals are computed by subtracting the fitted values from the original values.
-  The estimated values from the model are centered on zero by subtracting, for example, the mean from the values.
-  The centered fitted values and residuals are each sorted from smallest to largest values.
-  Side-by-side plots are created by plotting the sorted fitted values and residuals to their respective f-values (i.e. their fractional contribution) 




